---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Context and motivation {background-color="#491106ff"}

## Marketized welfare and its moral underpinnings

:::: {.incremental style="font-size: 120%;"}
- Commodification reshaped *who deserves what, and why*
- Policy feedback → rising **market justice preferences**: income-based access to health, education, pensions
- Moral economy: deservingness criteria mediate how structure shapes attitudes
- **Chile**: deep commodification; individual responsibility institutionalized across all three domains
::::

## Research problem

:::: {.incremental style="font-size: 120%;"}
- NICER framework → focus on the **effort** dimension of deservingness, operationalized as meritocratic beliefs
- Most work uses a single, summative notion of "meritocracy"
- This masks heterogeneity along **two crossed distinctions**:
  - meritocratic vs. **non-meritocratic** (privilege) beliefs
  - **perceptions** (what *is*) vs. **preferences** (what *ought to be*)
- Largely sidelined: **non-meritocratic** beliefs — privilege as a route to success
:::: 

## Multidimensional framework of meritocracy

::: {.fragment}

```{r}
#| fig-cap: "Castillo et al. (2023) measurement model"
#| fig-align: center
#| out-width: '100%'
#| fig-cap-location: bottom
library(knitr)
library(here)

knitr::include_graphics("images/diagramas(4).png")

```

:::

::: {.notes}

To approach meritocracy, we use a multidimensional framework. This model expands the usual way meritocracy is measured in empirical research.

:::
## Multidimensional framework of meritocracy



```{r}
#| fig-cap-location: bottom
#| fig-cap: "Castillo et al. (2023) measurement model"
#| fig-align: center
#| out-width: '100%'


knitr::include_graphics("images/diagramas(5).png")

```

::: {.notes}

Most studies focus only on meritocratic perceptions, that is, whether people believe that effort and ability are actually rewarded in society. This is represented in the upper-left corner of the diagram. 

 

:::

## Multidimensional framework of meritocracy


```{r}
#| fig-cap-location: bottom
#| fig-cap: "Castillo et al. (2023) measurement model"
#| fig-align: center
#| out-width: '100%'


knitr::include_graphics("images/diagramas(6).png")
```

::: {.notes}

What this model adds is, first, a distinction between perceptions and preferences, where perceptions refer to how people think society works and preferences refer to how they think it should work. This is represented in the upper-right corner using the same items, but with a normative sense. 


:::

## Multidimensional framework of meritocracy


```{r}
#| fig-cap-location: bottom
#| fig-cap: "Castillo et al. (2023) measurement model"
#| fig-align: center
#| out-width: '100%'


knitr::include_graphics("images/diagramas(7).png")
```


::: {.notes}

Second, it also incorporates non-meritocratic elements, captured through items on family wealth and social connections, both of which are shown at the bottom of the diagram. We call this dimension non-meritocratic privilege.

:::

## Multidimensional framework of meritocracy


```{r}
#| fig-cap-location: bottom
#| fig-cap: "Castillo et al. (2023) measurement model"
#| fig-align: center
#| out-width: '100%'


knitr::include_graphics("images/diagramas(8).png")
```

::: {.notes}
So, in total, the scale distinguishes four dimensions: perceived meritocracy, perceived non-meritocracy, preference for meritocracy, and preference for non-meritocracy. This means that recognizing privilege does not oppose supporting merit. Both can coexist within the same belief system in terms of perceptions and preferences.


Also, you can see the two items for each factor, all measured on a four-point agreement scale.

:::


## Research question & hypotheses

:::: {.fragment style="font-size: 120%;"}

**RQ:** How do meritocratic and privilege perceptions and preferences shape support for market-based welfare (health, education, pensions) in Chile?

::::


:::: {.incremental style="font-size: 120%;"}
- **H1**: Meritocratic *perceptions* → (+) market justice
- **H2**: Meritocratic *preferences* → (+) market justice
- **H3**: Privilege *perceptions*  → (–) market justice
- **H4**: Privilege *acceptance* → *open / exploratory*
:::: 